---
title: "Patel Glioblastoma- Dimension Reduction"
author: "Will Townes"
date: "5/20/2017"
output: html_document
---

```{r setup, include=FALSE}
library(modules) #devtools::install_github("klmr/modules")
reshape2<-import_package("reshape2")
bcse<-import_package("SummarizedExperiment")
import_package("Matrix",attach=TRUE)
import_package("ggplot2",attach=TRUE)
existing<-import("../algs/existing") #pca, zifa, etc
vamf<-import("../algs/vamf_stan") #our algorithm
cls<-import("../util/classifiers") #used for LDA, QDA, visualizations
fns<-import("../util/functions")
plx2name<-function(plx){paste0("tSNE_plx_",plx)}
```

# MGH26 Only

This is the only tumor spread across two batches so we use it to check for how each method does at removing the batch effect.

## Loading Data

First calculate size factors using scran

```{r}
#compute size factors with SCRAN (R 3.3+ only)
#library(scran)
load("data/MGH26/02_counts_genes_filtered.RData") #summarizedExperiment, includes a bulk sample!
meta<-bcse$colData(mgh26)
m<-as.matrix(bcse$assay(mgh26,"counts")[,meta$sampleType=="SC"])
meta<-meta[meta$sampleType=="SC",]
filter_genes<-function(m){m[rowSums(m>0)>3 & rowMeans(m)>0.2,]}
m<-filter_genes(m)
sizefactors <- scran::computeSumFactors(m)#,positive=TRUE)
library_size <- colSums(m)
plot(log(library_size),log(sizefactors)) #corr=.98
hist(sizefactors,breaks=20)
#write.table(sizefactors,file="data/MGH26/scran_sizefactors.txt",row.names=FALSE,col.names=FALSE)

meta$instrument<-as.factor(meta$instrument)
m<-Matrix(m)
d<-list(counts=m)
#raw counts needed for ZINB-WAVE
Y_raw<-matrix(as.vector(d$counts,mode="integer"),nrow=nrow(d$counts))
colnames(Y_raw)<-colnames(d$counts)
rownames(Y_raw)<-rownames(d$counts)
d$scran<-t(t(d$counts)/sizefactors)
Y<-Matrix(d$scran)
Y[Y<1]<-0 #detection threshold set higher than zero for "shot noise"
meta$detection_rate<-Matrix::colMeans(Y>0)
#meta$detection_rate<-meta$detection/nrow(d$counts)
meta<-as.data.frame(meta)
```

SCRAN size factors are highly correlated with library size.

## Visualization of Batch Effects

Detection Rates by Batch

```{r}
ggplot(meta,aes(x=instrument,y=detection_rate))+geom_boxplot()+theme_bw()+ylab("detection rate")
ggsave(file="results/plots/MGH26/detection_batch.pdf",height=6,width=8)
```

### Standard Dimension Reduction Methods

```{r}
img_save<-TRUE
patel_plt<-function(factors,meta){
  #x is data frame with factors, meta is data frame with batch info
  #batch name is "instrument"
  ggplot(cbind(meta,factors),aes(x=dim1,y=dim2,colour=instrument))+geom_point(size=3)+theme_bw()
}
plot_path<-file.path("./results/plots/MGH26/scran")
ggs<-function(plt,w=6,h=4){if(img_save) ggsave(file=file.path(plot_path,plt),width=w,height=h)}
```

```{r}
set.seed(101) #reproducible subsetting of dataset
g2k<-sample.int(nrow(Y),2000)
factors<-existing$pca(log2(1+Y[g2k,]))
patel_plt(factors,meta)
pd<-cbind(factors,det_rate=meta$detection_rate,instrument=meta$instrument)
round(cor(pd[,-4]),2) #.72 correlation
ggplot(pd,aes(x=det_rate,y=dim1,colour=instrument,shape=instrument))+geom_point(size=3)+theme_bw()+theme(legend.position="element_blank")+xlab("detection rate")+ylab("PC1")
ggs("pc1_vs_detrate_g2k.pdf")

write.table(pd,"../biostat_paper/data/patel/mgh26_pca_detrate_g2k.tsv",quot=FALSE,row.names=FALSE)
```

```{r}
set.seed(101) #reproducible subsetting of dataset
g2k<-sample.int(nrow(Y),2000)
Y_obs<-Y[g2k,]
dpath<-"data/MGH26/batch_separability"
fns$sparse2disk(Y_obs,file.path(dpath,"2Kgenes.tsv"))
Y_obs_l2<-log2(1+Y_obs)
#fns$sparse2disk(Y_obs_l2,file.path(dpath,"2Kgenes_log.tsv"))

### tSNE
dat<-list()
pplx<-20
dat[[plx2name(pplx)]]<-factors<-existing$tsne(Y_obs_l2,2,perplexity=pplx)
#fit<-cls$fit_classifier(factors,meta$instrument,"qda")
#cls$plot_classifier(fit,factors,meta$instrument)
patel_plt(factors,meta)+ggtitle(paste0("tSNE (perplexity ",pplx,")"))
ggs("tsne_plx_hi.pdf")
pplx<-5
dat[[plx2name(pplx)]]<-factors<-existing$tsne(Y_obs_l2,2,perplexity=pplx)
patel_plt(factors,meta)+ggtitle(paste0("tSNE (perplexity ",pplx,")"))
ggs("tsne_plx_lo.pdf")

### PCA
factors<-existing$pca(Y_obs_l2,3)
dat[["PCA"]]<-factors[,1:2]
patel_plt(factors,meta)+ggtitle("Principal Components Analysis")
ggs("pca12.pdf")
ggplot(cbind(factors,meta),aes(x=dim2,y=dim3,color=instrument))+geom_point(size=3)+theme_bw()+ggtitle("Principal Components Analysis")
ggs("pca23.pdf")

### MDS
dat[["MDS"]]<-existing$mds(Y_obs_l2,2,distance="manhattan")
patel_plt(dat[["MDS"]],meta)+ggtitle("Manhattan Multidimensional Scaling")
ggs("mds.pdf")
```

### ZINB-WAVE

```{r}
#library(zinbwave)
#zinb-wave uses raw counts, no normalization
system.time(zinb_res<-zinbwave::zinbFit(Y_raw[g2k,], K=2)) #100 seconds
#Warning message:
#In simpute.als(x, J, thresh, lambda, maxit, trace.it,  ... :
#  Convergence not achieved by 100 iterations
dat[["ZINB-WAVE"]]<-as.data.frame(zinb_res@W)
colnames(dat[["ZINB-WAVE"]])<-paste0("dim",1:2)
patel_plt(dat[["ZINB-WAVE"]],meta)+ggtitle("ZINB-WAVE")
ggs("zinbwave.pdf")
```

### ZIFA

```
time python ../util/zifa_wrapper.py data/MGH26/batch_separability/2Kgenes.tsv
```
7.5min
```{r}
factors <- read.table(file.path(dpath,"2Kgenes_zifa.tsv"))
colnames(factors)<-paste0("dim",1:2)
dat[["ZIFA"]]<-factors
patel_plt(factors,meta)+ggtitle("Zero Inflated Factor Analysis")
ggs("zifa.pdf")
```

### fSCLVM

```
time python ../util/fsclvm_wrapper.py data/MGH26/batch_separability/2Kgenes.tsv --nHiddenSparse 2
```
16sec
```{r}
factors <- read.table(file.path(dpath,"2Kgenes_fsclvm.tsv"))
colnames(factors)<-paste0("dim",1:2)
dat[["fSCLVM"]]<-factors
patel_plt(factors,meta)+ggtitle("f-SCLVM")
ggs("fsclvm.pdf")
```

### Varying-Censoring Aware Matrix Factorization

```{r}
system.time(res<-vamf$vamf(Y_obs,2)) #289sec, about 5min
#distribution of b1
hist(res$b1,probability=TRUE,breaks=20,xlim=c(0,1.2))
curve(dnorm(x,.75,.2),from=0,to=2,add=TRUE)
curve(dnorm(x,.5,.1),from=0,to=2,add=TRUE,lty=2)

dat[["VAMF"]]<-res$factors
patel_plt(res$factors,meta)+ggtitle("Censored Matrix Factorization")
ggs("vamf_mnar.pdf")
```
save b0, b1 scatter plot data for use in biostat paper
```{r}
vpd<-cbind(meta,b0=res$b0,b1=res$b1)
ggplot(vpd,aes(x=b0,y=b1,colour=instrument,shape=instrument))+geom_point()+theme_classic()
write.table(vpd,file="../biostat_paper/data/patel/mgh26_2Kgenes_vamf.tsv",row.names=FALSE)
```

### Comparing all methods

```{r}
pd<-do.call("rbind",lapply(names(dat),function(x){cbind(method=x,batch=meta$instrument,dat[[x]])}))
write.table(pd,file=file.path(dpath,"2Kgenes_all_methods.tsv"),row.names=FALSE)
#pd<-read.table(file.path(dpath,"2Kgenes_all_methods.tsv"),header=TRUE)
#pd$method<-factor(pd$method,levels=c("tSNE_plx_20","tSNE_plx_5","PCA","MDS","ZIFA","VAMF","fSCLVM","ZINB-WAVE"))
ggplot(subset(pd,!(method %in% c("MDS","ZINB-WAVE","fSCLVM"))), aes(x=dim1,y=dim2,colour=batch,shape=batch)) + geom_point() + facet_wrap(~method,scales="free",nrow=3)+theme_bw()+theme(legend.position="bottom")
ggs("all_methods.pdf",w=8,h=10.5)
#ggplot(pd,aes(x=dim1,y=dim2,colour=batch,shape=batch)) + geom_point() + facet_wrap(~method,scales="free",nrow=3)+theme_bw()+theme(legend.position="bottom")
```

save intermediate data for biostats paper

```{r}
write.table(pd,file="../biostat_paper/data/patel/mgh26_2Kgenes_all_methods.tsv",row.names=FALSE)
```

### Quantifying Separability over multiple random gene subsets

write replicate subsets to disk

```{r}
set.seed(101) #replicable subsetting of dataset
gsets<-replicate(100,sample.int(nrow(Y),2000),simplify=FALSE)
Ysubsets<-lapply(gsets,function(x){Y[x,]})
Ysubsets_raw<-lapply(gsets,function(x){Y_raw[x,]})

fpath<-file.path("data/MGH26/batch_separability/subsets_scran")
dir.create(fpath)
invisible(mapply(function(y,f){fns$sparse2disk(y,file.path(fpath,paste0(f,".tsv")))}, Ysubsets, 1:100))
fpath<-file.path("data/MGH26/batch_separability/subsets_raw")
dir.create(fpath)
for(i in 1:100){
  write.table(Ysubsets_raw[[i]],file.path(fpath,paste0(i,".txt")),quot=FALSE)
}
```

analyze replicates using VAMF script

```
seq 1 100 | parallel Rscript ../algs/vamf_script.R -p 1 -d 10 -i data/MGH26/batch_separability/subsets_scran/{}.tsv -o data/MGH26/batch_separability/vamf_out/{}.tsv
```

analyze replicates using ZIFA.
```
seq 1 100 | parallel python ../util/zifa_wrapper.py data/MGH26/batch_separability/subsets_scran/{}.tsv -o data/MGH26/batch_separability/zifa_out/{}.tsv
```

analyze replicates using ZINB-WAVE.
```
seq 1 100 | parallel Rscript ../algs/zinbwave_script.R -i data/MGH26/batch_separability/subsets_raw/{}.txt -o data/MGH26/batch_separability/zinbwave_out/{}.tsv
```

analyze replicates using fSCLVM.
```
seq 1 100 | parallel python ../util/fsclvm_wrapper.py data/MGH26/batch_separability/subsets_scran/{}.tsv -o data/MGH26/batch_separability/fsclvm_out/{}.tsv --nHiddenSparse 2
```

```{r}
nsim<-100
dat<-list()
for(plx in c(20,5)){
  dat[[plx2name(plx)]]<-lapply(Ysubsets,existing$tsne,perplexity=plx)
}
dat[["PCA"]]<-lapply(Ysubsets,existing$pca)
dat[["MDS"]]<-lapply(Ysubsets,existing$mds,distance="manhattan")
#read in ZIFA and VAMF results
fpath<-file.path("data/MGH26/batch_separability")
tmp_fn<-function(x,folder=c("vamf_out","zifa_out","zinbwave_out","fsclvm_out")){
  folder<-match.arg(folder)
  d<-file.path(fpath,folder,paste0(x,".tsv"))
  if(folder %in% c("zifa_out","fsclvm_out")){ #no column headers
    res<-read.table(d)
    colnames(res)<-paste0("dim",1:ncol(res))
    return(res)
  } else { #VAMF and ZINB-WAVE have column headers
    return(read.table(d,header=TRUE))
  }
}
dat[["ZIFA"]]<-lapply(seq_len(nsim),tmp_fn,folder="zifa_out")
dat_vamf<-lapply(seq_len(nsim),tmp_fn,folder="vamf_out")
dat[["VAMF"]]<-lapply(dat_vamf,function(x){x[,1:2]}) #restrict to top two dimension for comparability
dat[["ZINB-WAVE"]]<-lapply(seq_len(nsim),tmp_fn,folder="zinbwave_out")
dat[["fSCLVM"]]<-lapply(seq_len(nsim),tmp_fn,folder="fsclvm_out")
res<-expand.grid(method=names(dat),classifier=c("LDA","QDA"))
cls_err<-matrix(NA,nrow=nrow(res),ncol=nsim)
cls_funcs<-list(LDA=cls$lda_wrap,QDA=cls$qda_wrap)
for(i in 1:nrow(res)){
  cfn<-cls_funcs[[res[i,"classifier"]]]
  d<-as.character(res[i,"method"])
  cls_err[i,]<-vapply(dat[[d]],cfn,FUN.VALUE=1.0,labels=meta$instrument)
}
res<-cbind(res,cls_err)
res<-reshape2$melt(res,c("method","classifier"),variable.name="sim",value.name="cls_err")
res$Batch_Separability<-1-res$cls_err
#cache data for changing plots later if needed
write.table(res,file="results/plots/MGH26/scran/separability_plot_data.tsv",row.names=FALSE,quote=FALSE)

#save for biostat paper
write.table(res,file="../biostat_paper/data/patel/mgh26_batch_separability.tsv",row.names=FALSE,quote=FALSE)
```

```{r}
#res<-read.table("results/plots/MGH26/scran/separability_plot_data.tsv",header=TRUE)
res$method<-reorder(res$method,res$Batch_Separability,function(x){-median(x)})

plt1_methods<-c("VAMF","tSNE_plx_5","tSNE_plx_20","PCA","ZIFA")
ggplot(res[res$method %in% plt1_methods,], aes(x=method,y=Batch_Separability,fill=classifier))+geom_boxplot()+theme_classic()+coord_flip()+theme(legend.position="top")+ylab("batch separability")
ggsave("results/plots/MGH26/scran/separability_boxplot.pdf",width=8,height=6)

plt2_methods<-c("VAMF","fSCLVM","ZINB-WAVE")
ggplot(res[res$method %in% plt2_methods,], aes(x=method,y=Batch_Separability,fill=classifier))+geom_boxplot()+theme_classic()+coord_flip()+theme(legend.position="top")+ylab("batch separability")
ggsave("results/plots/MGH26/scran/separability_boxplot2.pdf",width=8,height=6)
```
