---
title: "Patel Glioblastoma- Dimension Reduction"
author: "Will Townes"
date: "5/20/2017"
output: html_document
---

```{r setup, include=FALSE}
library(modules) #devtools::install_github("klmr/modules")
reshape2<-import_package("reshape2")
bcse<-import_package("SummarizedExperiment")
import_package("Matrix",attach=TRUE)
import_package("ggplot2",attach=TRUE)
existing<-import("../algs/existing") #pca, zifa, etc
vamf<-import("../algs/vamf_stan") #our algorithm
cls<-import("../util/classifiers") #used for LDA, QDA, visualizations
fns<-import("../util/functions")
plx2name<-function(plx){paste0("tSNE_plx_",plx)}
```

# All Tumors

Exclude bulk RNA-Seq samples, non-tumor samples (eg cell lines), and samples not included by the authors in their analysis

## Loading Data

Filter for Patel's genes and cells in tumors, get sizefactors.

```{r}
#library(scran)
dir.create("data/all_tumors")
load("data/01_counts_genes.RData") #patel summarizedexperiment
gmeta<-scan(file="extdata/genes_patel_used.txt",what="character")
meta<-as.data.frame(bcse$colData(patel))
incl<-with(meta,includeSample=="TRUE" & cellType=="Glioblastoma" & sampleType=="SC")
patel_sc<-patel[rownames(patel) %in% gmeta, incl]
m<-as.matrix(bcse$assay(patel_sc))
meta<-as.data.frame(bcse$colData(patel_sc))

sizefactors <- scran::computeSumFactors(m)#,positive=TRUE)
library_size <- colSums(m)
plot(log(library_size),log(sizefactors)) #corr=.98
hist(sizefactors,breaks=20)

m<-Matrix(m)

meta$instrument<-as.factor(meta$instrument)
meta$tumorName<-as.factor(meta$tumorName)
cn<-colnames(meta)
cn[cn=="tumorName"]<-"Tumor"
cn[cn=="instrument"]<-"Instrument"
colnames(meta)<-cn
d<-list(counts=m)
d$scran<-t(t(d$counts)/sizefactors)
Y<-Matrix(d$scran)
Y[Y<1]<-0 #detection threshold higher than zero for 'shot noise'
fns$sparse2disk(Y,"data/all_tumors/Y.tsv")

#raw counts needed for ZINB-WAVE
Y_raw<-matrix(as.vector(d$counts,mode="integer"),nrow=nrow(d$counts))
colnames(Y_raw)<-colnames(d$counts)
rownames(Y_raw)<-rownames(d$counts)
write.table(Y_raw,"data/all_tumors/Y_raw.txt",quot=FALSE)
```

## Dimension Reduction Methods

Analyze using VAMF script. 5685 rows, 430 cols, latent dimension=10, timing was 35 minutes on laptop for 3 restarts.

```
time Rscript ../algs/vamf_script.R -d 10 -p 3 -i data/all_tumors/Y.tsv -o data/all_tumors/Y_vamf.tsv
```

Analyze using ZIFA. Took about 80 minutes (dimensionality=2)

```
time python ../util/zifa_wrapper.py data/all_tumors/Y.tsv -o data/all_tumors/Y_zifa.tsv
```

Analyze using ZINB-WAVE. 8 minutes (dim=2)

```
time Rscript ../algs/zinbwave_script.R -i data/all_tumors/Y_raw.txt -o data/all_tumors/Y_zinbwave.tsv
```

Analyze with fSCLVM. 2m48sec on laptop

```
time python ../util/fsclvm_wrapper.py data/all_tumors/Y.tsv -o data/all_tumors/Y_fsclvm.tsv --nHiddenSparse 2
```

```{r}
img_save<-TRUE

patel_plt<-function(factors,meta){
  #x is data frame with factors, meta is data frame with batch info
  #batch name is "instrument"
  ggplot(cbind(meta,factors),aes(x=dim1,y=dim2,colour=Tumor,shape=Instrument))+geom_point(size=2)+theme_bw()+labs(x="Dimension 1",y="Dimension 2")
}

Y_obs<-log2(1+Y)

### Detection Rates by Batch
meta$Detection_Rate<-Matrix::colMeans(Y_obs>0)
ggplot(meta,aes(x=Tumor,y=Detection_Rate,fill=Instrument))+geom_boxplot()+ylim(c(0,1))+ggtitle("Patel et al Detection Rates by Batch (all tumors)")+labs(y="Detection Rate")+theme_bw()
ggsave(file="results/plots/all_tumors/detection_batch.pdf",height=6,width=8)

plot_path<-file.path("./results/plots/all_tumors/scran")
dir.create(plot_path)
ggs<-function(plt,w=8,h=6){if(img_save) ggsave(file=file.path(plot_path,plt),width=w,height=h)}
```

```{r}
### tSNE
dat<-list()
pplx<-20
system.time(dat[[plx2name(pplx)]]<-factors<-existing$tsne(Y_obs,2,perplexity=pplx))
#ggplot(cbind(factors,meta),aes(x=dim1,y=dim2,colour=tumorName,shape=instrument))+geom_point(size=2)
patel_plt(factors,meta)+ggtitle(paste0("tSNE (perplexity ",pplx,")"))
ggs("tsne_plx_hi.pdf")
pplx<-5
system.time(dat[[plx2name(pplx)]]<-factors<-existing$tsne(Y_obs,2,perplexity=pplx))
patel_plt(factors,meta)+ggtitle(paste0("tSNE (perplexity ",pplx,")"))
ggs("tsne_plx_lo.pdf")

### PCA
system.time(factors<-existing$pca(Y_obs,3,scale=FALSE))
dat[["PCA"]]<-factors[,1:2]
patel_plt(factors,meta)+ggtitle("Principal Components Analysis")
ggs("pca12.pdf")
ggplot(cbind(factors,meta),aes(x=dim2,y=dim3,color=Tumor,shape=Instrument))+geom_point(size=2)+theme_bw()+ggtitle("Principal Components Analysis")+labs(x="Dimension 2",y="Dimension 3")
ggs("pca23.pdf")

### MDS
dat[["MDS"]]<-existing$mds(Y_obs,2,distance="manhattan",scale=FALSE)
patel_plt(dat[["MDS"]],meta)+ggtitle("Manhattan Multidimensional Scaling")
ggs("mds.pdf")

### ZIFA
factors <- read.table(file.path("data/all_tumors/Y_zifa.tsv"))
colnames(factors)<-paste0("dim",1:2)
dat[["ZIFA"]]<-factors
patel_plt(factors,meta)+ggtitle("Zero Inflated Factor Analysis")
ggs("zifa.pdf")

### ZINB-WAVE
factors<-read.table(file.path("data/all_tumors/Y_zinbwave.tsv"),header=TRUE)
dat[["ZINB-WAVE"]]<-factors
patel_plt(factors[,1:2],meta)+ggtitle("ZINB-WAVE")
ggs("zinbwave.pdf")

### fSCLVM
factors<-read.table(file.path("data/all_tumors/Y_fsclvm.tsv"))
colnames(factors)<-paste0("dim",1:2)
dat[["fSCLVM"]]<-factors
patel_plt(factors[,1:2],meta)+ggtitle("fSCLVM")
ggs("fsclvm.pdf")

### VAMF
factors<-read.table(file.path("data/all_tumors/Y_vamf.tsv"),header=TRUE)
barplot(vamf$colNorms(factors))
dat[["VAMF"]]<-factors[,1:2]
patel_plt(factors[,1:2],meta)+ggtitle("Varying-Censoring Aware Matrix Factorization")
ggs("vamf_mnar.pdf")
```

#### Comparing all methods

```{r}
plot_path<-file.path("./results/plots/all_tumors/scran")
pd<-do.call("rbind",lapply(names(dat),function(x){cbind(method=x,meta[,c("Instrument","Detection_Rate","Tumor")],dat[[x]])}))
write.table(pd,file=file.path(plot_path,"all_methods.tsv"),row.names=FALSE)

#cache for biostat paper
write.table(pd,file="../biostat_paper/data/patel/all_tumors_all_methods.tsv",row.names=FALSE,quote=FALSE)

#pd<-read.table(file.path(plot_path,"all_methods.tsv"),header=TRUE)
pd$method<-factor(pd$method)
plt1_methods<-c("VAMF","tSNE_plx_5","tSNE_plx_20","PCA","ZIFA")
ggplot(pd[pd$method %in% plt1_methods,], aes(x=dim1,y=dim2,colour=Tumor,shape=Instrument)) + geom_point() + facet_wrap(~method,scales="free",nrow=3)+theme_bw()+labs(x="Dimension 1",y="Dimension 2") #+ theme(legend.position="bottom")
ggs("all_methods1.pdf",w=8,h=10.5)

pd$Batch<-as.character(pd$Tumor)
tmp<-pd$Batch=="MGH26"
pd$Batch[tmp]<-paste(pd$Batch[tmp],pd$Instrument[tmp],sep="-")
pd$Batch<-as.factor(pd$Batch)
ggplot(pd[pd$method %in% plt1_methods,], aes(x=dim1,y=dim2,colour=Batch)) + geom_point(size=0.7) + facet_wrap(~method,scales="free",nrow=3)+theme_bw()+labs(x="Dimension 1",y="Dimension 2") #+ theme(legend.position="bottom")
ggs("all_methods2.pdf",w=8,h=10.5)

#separation of MGH26 batches?
ggplot(subset(pd,Tumor=="MGH26"),aes(x=dim1,y=dim2,colour=Batch))+geom_point(alpha=.5,size=.5)+facet_wrap(~method,scales="free")+theme_classic()
```

Correlations between PCs and detection rates

```{r}
ggplot(pd,aes(x=Detection_Rate,y=dim1,colour=Tumor,shape=Instrument))+geom_point(size=0.7)+facet_wrap(~method,scales="free",nrow=3)+theme_bw()+labs(x="Detection Rate",y="Dimension 1")
ggs("all_methods_detrate.pdf",w=8,h=10.5)
corvals<-plyr::ddply(pd,"method",plyr::summarise,corval=abs(cor(dim1,Detection_Rate)))
corvals<-corvals[order(corvals$corval),]
corvals$method<-factor(corvals$method,levels=levels(corvals$method)[order(corvals$corval)])
ggplot(corvals,aes(x=method,y=corval))+geom_bar(stat="identity")+theme_classic()
```