---
title: "Testing the Algorithm with Simulations"
author: "Will Townes"
date: "May 20, 2016"
output: html_document
---

## Setup

```{r}
options(mc.cores=3)
library(modules) #devtools::install_github('klmr/modules')
import_package("ggplot2",attach=TRUE)
Matrix<-import_package("Matrix")
sims<-import("../simulations/sims") #simulate missing data mechanisms
existing<-import("../algs/existing") #pca, zifa, etc
vamf<-import("../algs/vamf_stan") #our algorithm
img_save<-FALSE
```

Misc. functions used elsewhere

```{r}
plt_noise<-function(factors,batch){
  #create ggplot object
  #batch should be same length as ncol(factors)
  ggplot(factors,aes(x=dim1,y=dim2,colour=batch,shape=batch))+geom_point(size=3)+theme_bw()+labs(x="Dimension 1",y="Dimension 2")
}
plt_clst<-function(factors,ref){
  #ref is a data frame with "batch" and "id" variables
  ggplot(cbind(factors,ref),aes(x=dim1,y=dim2,colour=id,shape=batch))+geom_point(size=3)+theme_bw()+labs(x="Dimension 1",y="Dimension 2")
}
plx2name<-function(plx){paste0("tSNE_plx_",plx)}
logit<-function(p){log(p)-log(1-p)}
```

## Scenario I- Noise Only

Generate data that is just random noise, but allow two batches with different informative dropout rates. Expect traditional methods to detect two clusters but VAMF to find nothing of interest. The missingness rate will be roughly 60%.

```{r}
plot_path<-file.path("results/plots/noise_only")
ggs<-function(plt,w=3,h=3){if(img_save) ggsave(file=file.path(plot_path,plt),width=w,height=h)}
```

```{r}
set.seed(101)
sim1<-sims$noise_only(160,500,50) #N,G,Gsignal
dat<-list(Truth=sim1$ref[,c("dim1","dim2")])
batch<-sim1$ref$batch
Y_obs<-sim1$Y_obs
(cens<-signif(sum(Y_obs==0)/prod(dim(Y_obs)),3)) #missingness rate
#visualize missingness mechanism
cens_rates_obs<-Matrix$colMeans(Y_obs==0)
plot(sim1$ref$cens_rates,cens_rates_obs,xlab="expected censoring rate",ylab="empirical censoring rate",main="censoring rates by cell")
abline(0,1)
ggplot(dat[["Truth"]],aes(x=dim1,y=dim2,colour=batch,shape=batch))+geom_point(size=3)+theme_bw()+ggtitle("Original Latent Space")+labs(x="Dimension 1",y="Dimension 2")
ggs("00_original_space.pdf")

#write sparse matrix to disk
#summary() converts to COO (tuple) format
write.table(Matrix$summary(Y_obs),file="plot_data/noise_only.tsv",row.names=FALSE,quot=FALSE)
#Y_obs2<-do.call(Matrix$sparseMatrix,read.table("plot_data/noise_only.tsv",header=TRUE))
#stopifnot(max(abs(Y_obs2-Y_obs))<1e-5) #check for errors
```

### CIDR

```{r}
sData <- cidr::scDataConstructor(2^as.matrix(Y_obs)-1)
sData <- cidr::scDataConstructor(as.matrix(Y_obs))
sData <- cidr::determineDropoutCandidates(sData)
sData <- cidr::wThreshold(sData)
sData <- cidr::scDissim(sData)
sData <- cidr::scPCA(sData)
sData <- cidr::nPC(sData)
factors<-as.data.frame(sData@PC[,c(1,2)])
colnames(factors)<-c("dim1","dim2")
plt_noise(factors,batch)
```

### Standard Dimension Reduction Methods

```{r}
### tSNE
pplx<-20
system.time(dat[[plx2name(pplx)]]<-factors<-existing$tsne(Y_obs,2,perplexity=pplx))
plt_noise(factors,batch)+ggtitle(paste0("tSNE (perplexity ",pplx,")"))
ggs("tsne_plx_hi.pdf")
pplx<-5
system.time(dat[[plx2name(pplx)]]<-factors<-existing$tsne(Y_obs,2,perplexity=pplx))
plt_noise(factors,batch)+ggtitle(paste0("tSNE (perplexity ",pplx,")"))
ggs("tsne_plx_lo.pdf")

### PCA
system.time(factors<-existing$pca(Y_obs,3))
dat[["PCA"]]<-factors[,1:2]
factors$detection_rate<-1-cens_rates_obs
ggplot(factors,aes(x=detection_rate,y=dim1,color=batch,shape=batch))+geom_point(size=3)+theme_bw()+ggtitle("PC1 correlates with cell-specific detection rate")+labs(x="Detection Rate",y="Dimension 1")
cor(factors[,"dim1"],cens_rates_obs)
ggs("pca1_det.pdf")

plt_noise(factors,batch)+ggtitle("Principal Components Analysis")
ggs("pca12.pdf")
ggplot(factors,aes(x=dim2,y=dim3,color=batch,shape=batch))+geom_point(size=3)+theme_bw()+ggtitle("Principal Components Analysis")+labs(x="Dimension 2",y="Dimension 3")
ggs("pca23.pdf")

### Multidimensional Scaling
dat[["MDS"]]<-existing$mds(Y_obs,2,distance="manhattan")
plt_noise(dat[["MDS"]],batch)+ggtitle("Manhattan Multidimensional Scaling")
ggs("mds.pdf")
```

### ZIFA

```
time python ../util/zifa_wrapper.py plot_data/noise_only.tsv --log-transform False
```
This takes about 47s

```{r}
factors <- read.table("plot_data/noise_only_zifa.tsv")
colnames(factors)<-paste0("dim",1:2)
dat[["ZIFA"]]<-factors
plt_noise(factors,batch)+ggtitle("Zero Inflated Factor Analysis")
ggs("zifa.pdf")
```

### fSCLVM

```
time python ../util/fsclvm_wrapper.py plot_data/noise_only.tsv --log-transform False --nHiddenSparse 2
```
This took 6s

```{r}
factors <- read.table("plot_data/noise_only_fsclvm.tsv")

colnames(factors)<-paste0("dim",1:2)
dat[["fSCLVM"]]<-factors
plt_noise(factors,batch)+ggtitle("f-SCLVM")
#ggs("fsclvm.pdf")
colSums(factors) #0,0
```
fSCLVM shrinks sparse factors to zero when they are noise only. 

### Varying-censoring Aware Matrix Factorization

This is our algorithm.

```{r}
#Y_obs<-do.call(Matrix$sparseMatrix,read.table("plot_data/noise_only.tsv",header=TRUE))
system.time(res<-vamf$vamf(Y_obs,10,stmod="vamf_simple", log2trans=FALSE)) 
#126s for full model, 90s for model with fixed dropout pars
#batch<-rep(c("high detection","low detection"),each=80)
plt_noise(res$factors,batch)
barplot(existing$colNorms(res$factors)) #dimension learning?

dat[["VAMF"]]<-res$factors[,1:2]
plt_noise(dat[["VAMF"]],batch)+ggtitle("Varying-censoring Aware Matrix Factorization")
ggs("vamf.pdf")
```

### Comparison of Methods

```{r}
pd<-do.call("rbind",lapply(names(dat),function(x){cbind(method=x,Batch=batch,dat[[x]])}))
write.table(pd,file="plot_data/noise_only_all_methods.tsv",row.names=FALSE)
#pd<-read.table("plot_data/noise_only_all_methods.tsv",header=TRUE)
pd$method<-factor(pd$method,levels=c("Truth","VAMF","tSNE_plx_20","tSNE_plx_5","PCA","MDS","ZIFA","fSCLVM"))

#save for biostat paper
dir.create("../biostat_paper/data/simulations")
write.table(pd,file="../biostat_paper/data/simulations/noise_only_all_methods.tsv",quot=FALSE,row.names=FALSE)

ggplot(subset(pd,!(method %in% c("MDS","fSCLVM"))), aes(x=dim1,y=dim2,colour=Batch,shape=Batch)) + geom_point() + facet_wrap(~method,scales="free",nrow=3)+theme_bw()+labs(x="Dimension 1",y="Dimension 2")+theme(legend.position="top")
ggs("all_methods.pdf",w=4,h=6)
```

## Scenario II- Latent Clusters

```{r}
plot_path<-file.path("results/plots/latent_clusters")
ggs<-function(plt,w=3,h=3){if(img_save) ggsave(file=file.path(plot_path,plt),width=w,height=h)}
```

```{r}
set.seed(101)
sim2<-sims$latent_clusters(160,1000,50)
dat<-list(Truth=sim2$ref[,c("dim1","dim2")])
ref<-sim2$ref[,c("id","batch")]
Y_obs<-sim2$Y_obs
cens_rates<-sim2$ref$cens_rates
(cens<-signif(Matrix$mean(Y_obs==0),3)) #missingness rate
#visualize missingness mechanism
cens_rates_obs<-Matrix$colMeans(Y_obs==0)
plot(cens_rates,cens_rates_obs,xlab="expected censoring rate",ylab="empirical censoring rate",main="censoring rates by cell")
abline(0,1)
plt_clst(dat[["Truth"]],ref)+ggtitle("Original Latent Space")
ggs("00_original_space.pdf")
#write sparse matrix to disk
#summary() converts to COO (tuple) format
write.table(summary(Y_obs),file="plot_data/latent_clusters.tsv",row.names=FALSE,quot=FALSE)
```

### Standard Dimension Reduction Methods

```{r}
### tSNE
pplx<-20
system.time(dat[[plx2name(pplx)]]<-factors<-existing$tsne(Y_obs,2,perplexity=pplx))
plt_clst(factors,ref)+ggtitle(paste0("tSNE (perplexity ",pplx,")"))
ggs("tsne_plx_hi.pdf")
pplx<-5
system.time(dat[[plx2name(pplx)]]<-factors<-existing$tsne(Y_obs,2,perplexity=pplx))
plt_clst(factors,ref)+ggtitle(paste0("tSNE (perplexity ",pplx,")"))
ggs("tsne_plx_lo.pdf")

### PCA
system.time(factors<-existing$pca(Y_obs,3))
dat[["PCA"]]<-factors[,1:2]
factors$detection_rate<-1-cens_rates_obs
ggplot(cbind(factors,ref),aes(x=detection_rate,y=dim1,color=id,shape=batch))+geom_point(size=3)+theme_bw()+ggtitle("PC1 correlates with cell-specific detection")+labs(x="Detection Rate",y="Dimension 1")
cor(factors[,"dim1"],cens_rates)
ggs("pca1_det.pdf")

plt_clst(factors,ref)+ggtitle("Principal Components Analysis")
ggs("pca12.pdf")
ggplot(cbind(factors,ref),aes(x=dim2,y=dim3,color=id,shape=batch))+geom_point(size=3)+theme_bw()+ggtitle("Principal Components Analysis")+labs(x="Dimension 2",y="Dimension 3")
ggs("pca23.pdf")

### Multidimensional Scaling
dat[["MDS"]]<-existing$mds(Y_obs,2,distance="manhattan")
plt_clst(dat[["MDS"]],ref)+ggtitle("Manhattan Multidimensional Scaling")
ggs("mds.pdf")
```

### Variable-censoring Aware Matrix Factorization

This is our new algorithm VAMF. We also compare the effect of svmult tuning parameter on dimension learning.

```{r}
#Y_obs<-do.call(Matrix$sparseMatrix,read.table("plot_data/latent_clusters.tsv",header=TRUE))
system.time(res<-vamf$vamf(Y_obs,10,"vamf_simple",log2trans=FALSE)) 
#256 sec for full model, 163 for simple model

plt_clst(res$factors[,1:2],ref)
barplot(existing$colNorms(res$factors))
#plt_clst(res$factors[,1:2],ref)
#barplot(existing$colNorms(res$factors))

dat[["VAMF"]]<-res$factors[,1:2]
plt_clst(dat[["VAMF"]],ref)+ggtitle("Varying-censoring Aware Matrix Factorization")
ggs("vamf.pdf")
```

VAMF also learns the correct dimensionality

```{r}
pdf("results/plots/latent_clusters/vamf_dim_learn.pdf",8,6)
barplot(existing$colNorms(res$factors),xlab="Latent Dimension",ylab="L2 Norm of Posterior Mean")#,main="VAMF Automatic Dimensionality Learning")
dev.off()
```

### ZIFA

```
time python ../util/zifa_wrapper.py plot_data/latent_clusters.tsv --log-transform False
```
3m25s

```{r}
factors <- read.table("plot_data/latent_clusters_zifa.tsv")
colnames(factors)<-paste0("dim",1:2)
dat[["ZIFA"]]<-factors
plt_clst(factors,ref)+ggtitle("Zero Inflated Factor Analysis")
ggs("zifa.pdf")
```

### fSCLVM

```
time python ../util/fsclvm_wrapper.py plot_data/latent_clusters.tsv --log-transform False --nHiddenSparse 2
```
This took 9 seconds

```{r}
factors <- read.table("plot_data/latent_clusters_fsclvm.tsv")
colnames(factors)<-paste0("dim",1:2)
dat[["fSCLVM"]]<-factors
plt_clst(factors,ref)+ggtitle("f-SCLVM")
ggs("fsclvm.pdf")
```
fSCLVM does a good job removing the detection rate bias by absorbing it into dense factors, but can be too aggressive at pruning away dimensions.

### SIMLR

```{r}
system.time(res<-SIMLR::SIMLR(as.matrix(Y_obs),4,2))
factors<-res$ydata
colnames(factors)<-c("dim1","dim2")
plt_clst(factors,ref)
```

### PPCA

```{r}
system.time(res<-pcaMethods::pca(t(as.matrix(Y_obs)),"ppca",2))
factors<-res@scores
colnames(factors)<-c("dim1","dim2")
plt_clst(factors,ref)
```

### CIDR

```{r}
#sData <- cidr::scDataConstructor(2^as.matrix(Y_obs)-1)
sData <- cidr::scDataConstructor(as.matrix(Y_obs))
sData <- cidr::determineDropoutCandidates(sData)
sData <- cidr::wThreshold(sData)
sData <- cidr::scDissim(sData)
sData <- cidr::scPCA(sData)
sData <- cidr::nPC(sData)
factors<-sData@PC[,c(1,2)]
colnames(factors)<-c("dim1","dim2")
plt_clst(factors,ref)
```

### Comparison of Methods

```{r}
pd<-do.call("rbind",lapply(names(dat),function(x){cbind(method=x,ref,dat[[x]])}))
write.table(pd,file="plot_data/latent_clusters_all_methods.tsv",row.names=FALSE)
#pd<-read.table("plot_data/latent_clusters_all_methods.tsv",header=TRUE)
pd$method<-factor(pd$method,levels=c("Truth","VAMF","tSNE_plx_20","tSNE_plx_5","PCA","MDS","ZIFA","fSCLVM"))
pd$id<-as.factor(pd$id)
colnames(pd)[2:3]<-c("Cluster_ID","Batch")

#save for biostat paper
write.table(pd,file="../biostat_paper/data/simulations/latent_clusters_all_methods.tsv",row.names=FALSE,quot=FALSE)

ggplot(subset(pd,!(method %in% c("MDS","fSCLVM"))), aes(x=dim1,y=dim2,colour=Cluster_ID,shape=Batch)) + geom_point() + facet_wrap(~method,scales="free",nrow=3)+theme_bw()+labs(x="Dimension 1",y="Dimension 2")+theme(legend.position="top")

ggs("all_methods.pdf",w=4,h=6)
```